---
title: "Data Ingestion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Ingestion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

## Overview

dbSequence stores genomic data in DuckDB, enabling analysis of datasets larger than available memory. Data stays in the database—R operates on lazy references.

## Supported Formats

| Format | Function | BiocIO Class |
|--------|----------|--------------|
| BED | `read_bed()` | `BEDFile` |
| VCF | `read_vcf()` | `VcfFile` |
| GFF/GTF | `read_gff()` | `GFFFile` |
| BAM | `read_bam()` | `BamFile` |

## Quick Start

```{r quick-start}
library(dbSequence)

# Import a BED file
db_seq <- read_bed("fragments.bed")

# Data stays in DuckDB - this is a lazy reference
db_seq
```

## BiocIO Pattern

For more control, use the BiocIO `import()` function:
```{r biocio-pattern}
library(BiocIO)
library(rtracklayer)

# Specify destination database
db_seq <- import(
    BEDFile("fragments.bed"),
    dest = DuckDBFile("my_database.duckdb"),
    table_name = "fragments"
)
```

## Lazy Evaluation

The key feature: data never loads into R memory.

```{r lazy-demo}
# This does NOT load data
db_seq <- read_bed("large_file.bed")

# Still no data in memory - just adds a filter condition
filtered <- dplyr::filter(db_seq@value, seqnames == "chr1")

# Data only loads when you explicitly collect
result <- dplyr::collect(filtered) # NOW data enters R
```

## Multiple Tables

Store multiple datasets in one database:
```{r multiple-tables}
db <- DuckDBFile("my_project.duckdb")

# Import different files to same database
peaks <- import(BEDFile("peaks.bed"), dest = db, table_name = "peaks")
genes <- import(GFFFile("genes.gff3"), dest = db, table_name = "genes")
```

## Connection Management

dbSequence manages connections automatically, but for long sessions:
```{r connections}
# Access the underlying connection
con <- dbplyr::remote_con(db_seq@value)

# For manual control, create your own connection
con <- DBI::dbConnect(duckdb::duckdb(), "my_database.duckdb")
# ... do work ...
DBI::dbDisconnect(con, shutdown = TRUE)
```

## Next Steps

- [plyranges API](plyranges-api.html) - Range operations on lazy data

