---
title: "plyranges API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{plyranges API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = TRUE,
    warning = FALSE,
    message = FALSE
)
library(dbSequence)
library(GenomicRanges)
library(dbplyr)
```

## Overview

dbSequence provides plyranges-compatible functions that translate to SQL queries. Operations execute in DuckDB, not R, enabling analysis of datasets larger than memory.

## Core Functions

| Function | Description | Returns |
|----------|-------------|---------|
| `filter_by_overlaps(x, y)` | Ranges in x overlapping y | `dbSequence` (lazy) |
| `compute_coverage(x, region, window)` | Binned coverage counts | `data.frame` |
| `as_granges(x)` | Convert to GRanges | `GRanges` (collected) |

## filter_by_overlaps

Find ranges overlapping a region of interest:

```{r filter-overlaps}
# Load example data
bed_file <- system.file("extdata", "example.bed", package = "dbSequence")
fragments <- read_bed(bed_file)

# Define region of interest (example.bed has features on chr1:100-200, 300-400)
region <- GRanges("chr1:100-500")

# Filter to overlapping fragments - stays lazy
overlapping <- filter_by_overlaps(fragments, region)

overlapping
```

### Multiple Regions

```{r multi-region}
# Filter against multiple peaks
peaks <- GRanges(c("chr1:100-150", "chr1:300-350"))
overlapping <- filter_by_overlaps(fragments, peaks)
```

### dbSequence to dbSequence

Both arguments can be lazy. To join them, they must reside in the same database and **share the same connection object**.

```{r lazy-join}
# Use a temporary persistent DB
temp_db <- tempfile(fileext = ".duckdb")
dest <- DuckDBFile(temp_db)

# Import both tables to the same database file
peaks_orig <- read_bed(bed_file, dest = dest, table_name = "peaks", lazy = FALSE)
frags_orig <- read_bed(bed_file, dest = dest, table_name = "fragments", lazy = FALSE)

# Create a shared connection (reuse the one from peaks_orig)
# This is required because dplyr queries cannot span multiple connections
con <- dbProject::conn(peaks_orig)
frags_shared <- dbSequence("fragments", file_source = dest, .conn = con)

# Now both objects share 'con', so we can join them lazily
overlapping <- filter_by_overlaps(frags_shared, peaks_orig)
```

## compute_coverage

Calculate binned coverage for a region:

```{r coverage}
# 50bp bins across region
coverage <- compute_coverage(
    fragments,
    region = "chr1:0-500",
    window = 50
)

# Returns data.frame with bin_start, bin_end, count
head(coverage)
```

## as_granges

Convert to GRanges when you need Bioconductor operations:

```{r as-granges}
# Collect subset to GRanges
gr <- fragments |>
    filter_by_overlaps(region) |>
    as_granges()

# Now use standard GenomicRanges operations
GenomicRanges::reduce(gr)
```

> **Note:** `as_granges()` collects data into R memory. Filter first.

## Workflow Example

Typical analysis pattern:
```{r workflow}
# 1. Load data (lazy) - using example.bed
fragments <- read_bed(bed_file)

# 2. Define analysis region
promoter <- GRanges("chr1:50-250")

# 3. Filter to region (still lazy)
promoter_fragments <- filter_by_overlaps(fragments, promoter)

# 4. Compute coverage
coverage <- compute_coverage(promoter_fragments, promoter, window = 10)

# 5. Collect if needed for custom analysis
gr <- as_granges(promoter_fragments)
```

## File Type Restrictions

BAM/CRAM files are blocked from `filter_by_overlaps` due to their indexed nature—use `Rsamtools` instead:

```{r bam-error, error=TRUE}
bam_file <- system.file("extdata", "example.bam", package = "dbSequence")
bam_data <- read_bam(bam_file)

# This should error
filter_by_overlaps(bam_data, region)
```

## Next Steps

```{r sessionInfo}
sessionInfo()
```
